Chain separation distribution in protein contact networks

### 2022-05-23 paper/main.tex

%\documentclass{article}
%\usepackage[a4paper, total={6in, 8in}]{geometry}
%\documentclass[reprint,amsmath,amssymb,rmp,onecolumn,notitlepage,11pt]{revtex4-1}
\documentclass[
reprint,
twocolumn,
%preprint
%onecolumn,
amsmath,amssymb,superscriptaddress,aps,
pre]{revtex4-1}
\usepackage[utf8]{inputenc}
%\usepackage{authblk}
%\usepackage{natbib}
\usepackage[normalem]{ulem}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{xcolor}
\newcommand{\red}[1]{\textcolor{red!80!black}{#1}}
\newcommand{\blue}[1]{\textcolor{blue!80!black}{#1}}
\newcommand{\green}[1]{\textcolor{green!70!black}{#1}}
\newcommand{\gray}[1]{\textcolor{gray!80!black}{#1}}
\usepackage{mathtools}
\DeclarePairedDelimiter{\evdel}{\langle}{\rangle}
\usepackage[colorinlistoftodos]{todonotes}
\newcommand{\Pin}{P_{\mathrm{in}}}
\newcommand{\fhigh}{f_{\mathrm{high}}}
\newcommand{\flow}{f_{\mathrm{low}}}

\begin{document}
\title{Chain separation distribution in protein contact networks}
\title{Theoretical approximation of chain separation distribution in protein contact networks}
\title{Theory of chain separation distributions gives insights in real protein contact networks}
%\title{}
\author{Nora Molkenthin}
\affiliation{Potsdam Institut für Klimafolgenforschung}
\affiliation{Center for Advancing Electronics Dresden (cfaed), Technical University of Dresden}
\author{J. J.  Güven}
\affiliation{EaStCHEM School of Chemistry, University of Edinburgh, Edinburgh, United Kingdom}
\author{Steffen Mühle}
\affiliation{Max-Planck Institute for Dynamics and Self-Organization, 37077 Göttingen, Germany}
\author{Antonia S J S Mey}
\email[Electronic Address: ]{antonia.mey@ed.ac.uk}
\affiliation{EaStCHEM School of Chemistry, University of Edinburgh, Edinburgh, United Kingdom}

\begin{abstract}
The mechanisms by which we can determine a protein's 3D structure based on the protein's amino acid sequence have long been one of the key mysteries of biophysics. With the advent of AlphaFold2, making predictions for protein structures based on amino acid sequences at an accuracy similar to experiments is now possible, yet requires a lot of available structural data for training the model. Predicting contact maps from sequence \textit{de novo} is an appealing alternative. Here, we are interested in understanding the emergent behavior of protein contact maps derived from a network ensemble perspective of a geometrically constrained random interaction model. We show that the frequency distributions of amino acid distances obtained from simulations of the geometrically constrained interaction model can be described by an analytical approximation of the two-dimensional analogue of this model. To this end, we look at the scaling behavior of the simulation with the analytical approximation for different chain lengths typically found in proteins with good agreement. Furthermore, we compare our constrained model with frequency distributions of amino acid distances from contact maps of real proteins---from the protein data bank and AlphaFold2 database---showing that these follow a similar trend as the geometrically constrained model.   
 We therefore suggest the conclusion that the approximate power-law behavior of the amino acid distance distribution can be attributed to geometric constraints of protein chains and amino acid sequences only play a secondary role.  
%The distribution of chain separations of interacting amino acids in proteins roughly follows a power law. Here we show, analytically and in simulations, that a geometrical stochastic model of folding chains can explain this behavior. 
\end{abstract}
\maketitle

\section{Introduction}
Proteins, the molecular machines of every living organism perform vital tasks required for life to persist, ranging from transport (e.g. hemoglobin)~\cite{ahmed2020hemoglobin}, signal transduction (e.g. rhodopsin)~\cite{nagata2021rhodopsins}, immune responses (e.g. antibodies), and hormonal regulation (e.g. insulin)~\cite{dill2008protein,Dill1042}. All natural proteins are made of 20 different amino acids which dictate the three-dimensional conformations the proteins adopt in order to function~\cite{scheraga2007protein}. One of the great challenges has been to understand how the primary structure, the amino acid sequence, can lead to the fully folded functional protein. The last 50 years has seen many different routes to computationally predict biologically active protein structures without having to solve a crystal structure of the protein. [input information from Rohan's presentation]. AlphaFold2, a machine learning model for structure prediction, has shown predictiveness at a level of experimental accuracy in the most recent protein structure prediction challenge~\cite{jumper2021highly, kryshtafovych2021critical}. This has lead to an explosion of available structural protein models beyond the protein database, containing structures from crystallographic, NMR, and cryo-EM experiments~\cite{berman2000protein}. Beyond the prediction of structure understanding emergent patterns in proteins on a more fundamental level is of interest, as it opens up a deeper understanding of proteins structure and function relationship.  One way of looking at the functional forms of proteins and classifying their structural properties is by looking at protein contact maps (PCM) or protein contact networks (PCN)~\cite{Vendruscolo2002,DiPaola2013,Estrada2011}. A PCM---we will use PCMs in favor of PCN in this paper---is the resulting adjacency matrix of spatially close amino acids, given a certain cut-off as determined typically from a protein structure. PCMs have shown promise in understanding protein folding patterns as well as revealing allosteric communication pathways~\cite{yao2019establishing, menichetti2016network,dokholyan2002topological}. Their use is also prominently featured in machine learning approaches. One challenge is to predict a PCM \textit{de novo} from sequence alone~\cite{bassot2019using, rives2021biological, rao2021msa}, without using structural data as part of the training set, as was done for AlphaFold2. These PCMs can then be used in e.g. graph convolutional machine learning approaches to train a model able to predict binding affinities for certain types of proteins~\cite{jiang2020drug},
From a physicist's perspective it is interesting to understand the emergent behavior in protein structure by looking at PCMs. Since a PCM is an adjacency matrix and therefore a graph, it is possible to investigate PCMs using common network measures. For example, Bartoli \textit{et al.}~\cite{bartoli2008effecta} build a model in which they assume that proteins have an amino acid distance distribution that follows $\frac{1}{s}$, where $s$ is the separation of two connected amino acids along the backbone chain (in the following called \emph{amino acid distance}) as the simplest implementation of a connection probability, thus decreasing with the distance along the sequence. However, they do not provide a quantitative explanation for this assumption.

Here, we show how a simple geometrical model can reproduce this heuristically found behavior well, and even provide a simple analytical approximation. The three-dimensional geometrical model and the analytical approximation of its two-dimensional analogue both build on the idea that inherently geometrical objects, such as amino acids, imply that any resulting PCM network ensemble modelling them has to be spatially embedded~\cite{molkenthin2016scaling, molkenthin2020self}. 
Different approaches have been used in the past for geometrical models describing protein folding, some of which derived characteristics of the secondary structure from constraints on bond and torsion angles~\cite{bhattacharjee2013flory,Danielsson2010,Molkenthin2011} and others focused on the formation mechanisms of the tertiary structure~\cite{molkenthin2016scaling, molkenthin2020self}.

Based on such considerations, the network formation process repeatedly connects random pairs of units, without causing any intersections along the chain~\cite{molkenthin2016scaling}. Such a model can now be used to investigate the emergent behavior of amino acid distance distributions as used heuristically by Bartoli \textit{et al.}~\cite{bartoli2008effecta}. To this end, we define and analyze the amino acid distance distributions $P(s)$ that emerge from the geometric model, where the \emph{amino acid distance} $s$ of a link is defined as the distance of the two end units along the original chain.

This distribution is then compared to the amino acid distance distribution extracted from protein contact networks generated from protein data bank data.

We find that, despite the highly simplified model, the generated amino acid distance distribution is in agreement with the one observed in networks generated from PDB data and roughly consistent with the heuristic $1/s$ finding. The analytical approximation provides a correction term on top of the $1/s$ heuristic.

%%%%%%%%%%%%%%%%%
% Section       %
%%%%%%%%%%%%%%%%%
\section{Protein contact maps}
The complex interaction pattern between the amino acids in a protein can be very naturally expressed as a network or graph, in which each amino acid is represented by a node and spatial proximity is encoded as a link. Whenever two central $C_\alpha$ atoms are closer together than a threshold $d_c$, they are connected, linked, or in `close contact'. These connections can be determined from the 3D functional or folded structure of the protein and are encoded in a so called contact map. This contact map is effectively an adjacency matrix $\textbf{A}^{\mathrm{struc}}$ or binary matrix
\begin{equation}
  A^{\mathrm{struc}}_{ij}=
  \begin{cases}
   0, & \text{ if } d_{i,j}>d_c \text{ or } i=j\\
      1, & \text{ if } d_{i,j}\leq d_c .
      \end{cases}
    \label{eq:aij}
\end{equation}
Adjacency matrices are commonly used to describe networks and make up a \textit{protein contact map}. 
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\columnwidth]{figures/Fig1/Fig1.pdf}
    \caption{a) The amino acid distance $s$ is defined as the separation along the chain (teal) of two connected units (pink). b)~The solid green link prohibits some connections from node~$A$ (dotted pink), while others are still permitted (dotted green). Counting all newly excluded links, we find that a link of length $s$ prohibits $2(l-1)$ links of length $l$ up to a maximum of $s$ and $2(s-1)$ links of length $s$ or more. c)~Example of a protein contact map taken from a 2D simulation. Pink squares indicate a contact is present between two amino acids.}
    \label{fig:schematic}
\end{figure}

@@

Based on the basic observation that amino acids are objects in space which cannot overlap indefinitely, we introduced a model in~\cite{molkenthin2020self} starting from a chain of identical spheres, each in contact only with the neighbours it is connected to. Then additional connections form by randomly selecting two spheres and moving them towards each other until they touch, the new links formed that way can not be broken in later steps and function as constraints on subsequent links. If contact of the two selected spheres is geometrically impossible without breaking previously made connections or leading to overlaps of any spheres, the link is not made and taken out of the pool of possible connections. This process is repeated until no more links can be formed without violating the geometric constraints and the final structure can be expressed as the adjacency matrix
\begin{equation}
  \mathbf{A}^{\mathrm{sim}}_{ij}=
  \begin{cases}
   1, & \text{ if }|i-j|=1 \text{ or i and j  randomly connected} \\
      0, & \text{ otherwise }  .
      \end{cases}
    \label{eq:aij}
\end{equation}

In~\cite{molkenthin2016scaling}, we have introduced a simplified, two-dimensional version of this model, which starts from a closed chain (i.e. a ring) of $N$ unit discs and subsequently adds links, such that connected discs touch, yet no discs intersect. The advantage of this model is that it can be approximated by a purely topological simplification, which can be treated analytically. Fig.~\ref{fig:schematic} illustrates the construction of the 2D geometric model.

The topological constraints in the resulting network model are:
(a) New links always form between two units that are part of the same face of the graph (region enclosed by a cycle in the network). This prevents the overlapping of discs. (b) No links form across the outer face. This prevents the enclosure of a unit by less than six other units (which is geometrically impossible) such that (c) the maximum degree of each unit is six, as six is the maximum number of unit discs one central unit disc can touch. (d) Once connected by a link, pairs of units do not disconnect.

%%%%Missing%%%%%
% TO DO: @Nora
% Describe how the protein contact maps are generated in the geometric constrained model %
% Let's call the adjacency matrix in this case \mathbf{A}^{\mathrm{sim}}_{ij} where sim stands for simulation. 
%%%%%%%%%%%%%%%

@@

The underlying formation mechanisms of individual protein folds are incredibly complex and depend on the surrounding solvent, as well as the specific amino acid sequence and their quantum-mechanical interactions. Here we want to assess if the geometrically constrained protein model's contact map distributions capture real protein behavior as suggested in~\cite{bartoli2008effecta}. To understand this better, we want to study the ensemble of PCMs from real protein structures. To this end we will look at `averaged' contact maps over many different proteins that have different amino acid sequences, but their overall sequence or chain length is the same. 
We will use the RCSB protein data bank (PDB), a database of structurally resolved protein amino acid sequences through X-ray crystallography, NMR or cryo-EM experiments and the synthetically generated structural data base from the machine learned structures of AlphaFold2~\cite{jumper2021highly}. The PDB contains just under 200,000 protein structures to date~\cite{PDB}. The distribution of amino acid chain lengths in the PDB is illustrated in Fig.~\ref{fig:pdb_stats} b. Protein chain lengths in the PDB vary from small fragments of less than 10 amino acids to large agglomerates of over 2000 amino acids. The most frequently occurring chain lengths, however, are between 100 and 300 amino acids as shown in Fig.~\ref{fig:pdb_stats} b. The pink subset of the barplot shown in Fig.~\ref{fig:pdb_stats} b, are the set of structures used that fulfill the criterion on amino acid chain lengths between 85 and 115 amino acids, 185 and 215 amino acids and 285 and 315 amino acids and represent all structures for the analysis. The second database consists of synthetically generated structures using AlphaFold2 with sequences taken from SwissProt~\cite{alpha fold 2 database}. SwissProt is a manually curated database of protein sequences containing over 500,000 protein sequences~\cite{SwissProt}, whose amino acid chain length distribution is shown in Fig.~\ref{fig:pdb_stats} a in blue-green. The pink bars again indicate the subset of AlphaFold2 structures used for the contact map analysis.

 \begin{figure}[t]
        \centering
	\includegraphics[width=0.8\columnwidth]{paper/figures/Fig2/Fig2.pdf}
	    \caption{Histogram of protein sequence lengths for protein structures from two structure databases. a) AlphaFold2 predicted protein structures ($\alpha$F) taken from all protein sequences in the reviewed Uniprot Knowledge Base (Swiss-Prot) database based on their Uniprot ids. Pink bars shows fraction of structures used for the average contact map analysis.  b) RSCB PDB chain length frequencies based on sequence length. Pink bars in plot show proteins structures whose sequence lengths matches the resolution of structure accounting for the whole sequence. Only pink PDB samples were used for the average contact map analysis. c) Example of a protein contact map taken from PDB ID: 1AKI, with a cut-off of 8~Å. Pink squares indicate a contact present between amino acids.}
        \label{fig:pdb_stats}
\end{figure}
All structure based contact maps are generated in the same way for AlphaFold2 structures and PDB structures. Structures used were downloaded in pdb format, read in with MDAnalysis version 2.0.0~\cite{gowers2016mdanalysis} and only C$_{\alpha}$ were selected and used for the analysis. Proteins were categorised by length and placed into three groups. Amino acid chain lengths between 85 and 115 amino acids, 185-215 amino acids and 285-315. For each of the used sequences, all pairwise C$_{\alpha}$ distances were assessed and two C$_{\alpha}$ atoms were in contact if their distance was less than 8~Å. All pairwise distances were evaluated and in this way, building up the binary contact map or adjacency $A_{ij}^{\mathrm{struc}}$.  
All code for the construction of the contact maps both for the geometrically constrained system and structural protein data and their analysis can be found on github at: \url{https://github.com/meyresearch/sequence_distance_distribution}.

%%%%%%%%%%%%
% Section. %
%%%%%%%%%%%%
\section{Analytic approximation for the geometrically constrained model}

To assess how we can best describe the distribution of the probability of amino acid distances as obtained by a PCM for the geometrically constrained model, we propose the following analytical approximation. To be able to do this we will look at the amino acid distance distribution for the simplified 2D model. Fig~\ref{fig:schematic} summarizes the idea behind this two-dimensional model.

Let us start by introducing the auxiliary variable $F_k(s)$, defined to be the number of possible links with an amino acid distance of $s$ if $k$ links have been added before. Before any links are added, all amino acid distances are equally likely, as there are $F_0(s)=N$ possibilities of making a link of each distance $s$ with $2\leq s < N/2$.

As we add more links, not only are links taken out of this pool, because they have already been realized, \red{but} each existing link can also geometrically prohibit other connections. Adding a link of length $s_1$ prohibits $2(s-1)$ links of length $s<s_1$ and $2(s_1-1)$ links of length $s>s_1$ (see Fig.~\ref{fig:schematic}~b). To find the expected number of links still available in subsequent steps, we average over all possible amino acid distances $s_1$ of the first link added.

The expected distribution of possible links after one step is \red{thus given} by the average over available link pools taken over all possible values of $s_1$
\begin{equation}
\begin{aligned}
    F_1(s)&=\frac{1}{N/2-1} \sum_{s_1=2}^{N/2}{ \begin{cases}
    F_0(s)-2(s-1) \text{ , for } s<s_1\\
    F_0(s)-2(s_1 -1)\text{ , for } s\geq s_1
    \end{cases}}
    \nonumber \\
    &\approx\frac{2}{N} \sum_{s_1=2}^{N/2} { \begin{cases}
    N-2(s-1) \text{ , for } s<s_1\\
    N-2(s_1 -1)\text{ , for } s\geq s_1
    \end{cases}}
\end{aligned}
    \label{eq:reduction_first_step}
\end{equation}

>>>>

%\documentclass{article}
%\usepackage[a4paper, total={6in, 8in}]{geometry}
%\documentclass[reprint,amsmath,amssymb,rmp,onecolumn,notitlepage,11pt]{revtex4-1}
\documentclass[
reprint,
twocolumn,
%preprint
%onecolumn,
amsmath,amssymb,superscriptaddress,aps,
pre]{revtex4-1}
\usepackage[utf8]{inputenc}
%\usepackage{authblk}
%\usepackage{natbib}
\usepackage[normalem]{ulem}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{xcolor}
\newcommand{\red}[1]{\textcolor{red!80!black}{#1}}
\newcommand{\blue}[1]{\textcolor{blue!80!black}{#1}}
\newcommand{\green}[1]{\textcolor{green!70!black}{#1}}
\newcommand{\gray}[1]{\textcolor{gray!80!black}{#1}}
\usepackage{mathtools}
\DeclarePairedDelimiter{\evdel}{\langle}{\rangle}
\usepackage[colorinlistoftodos]{todonotes}
\newcommand{\Pin}{P_{\mathrm{next}}}
\newcommand{\fhigh}{f_{\mathrm{high}}}
\newcommand{\flow}{f_{\mathrm{low}}}

\begin{document}
\title{Chain separation distribution in protein contact networks}
\title{Theoretical approximation of chain separation distribution in protein contact networks}
\title{What geometrically constrained protein models can tell us about real-world protein contact maps}
%\title{}
\author{Nora Molkenthin}
\email[Author to whom correspondence should be addressed:]{nora.molkenthin@pik-potsdam.de}
\affiliation{Potsdam Institute for Climate Impact Research, Germany}
\author{J. J. Güven}
\affiliation{EaStCHEM School of Chemistry, University of Edinburgh, Edinburgh, United Kingdom}
\author{Steffen Mühle}
\affiliation{Max-Planck Institute for Dynamics and Self-Organization (MPIDS), Am Faßberg 17, 37077 Göttingen, Germany}
\author{Antonia S J S Mey}
\email[Author to whom correspondence should be addressed:]{antonia.mey@ed.ac.uk}
\affiliation{EaStCHEM School of Chemistry, University of Edinburgh, Edinburgh, United Kingdom}

\begin{abstract}
The mechanisms by which a protein's 3D structure can be determined based on its amino acid sequence have long been one of the key mysteries of biophysics. Often simplistic models, e.g. derived from geometric constraints, capture bulk real-world 3D protein--protein properties well. One approach is using protein contact maps to better understand protein's properties. Here, we investigate the emergent behavior of protein contact maps derived from a geometrically constrained random interaction model in comparison to real-world proteins. Deriving an analytical approximation for the distribution of amino acid distances $s$ of a 2D geometric model, by means of a mean-field approach, we present a theoretical basis for the distribution of amino acid distances in proteins. To validate this approximation we compare it to simulations of a 2D and 3D geometrically constrained model, as well as contact maps of real proteins. Using data from the RCSB Protein Data Bank (PDB) and AlphaFold~2 database, the analytical approximation can be accurately fit to both datasets. This holds for protein chain lengths of $L\approx100$, $L\approx200$, and $L\approx300$. While, a universal scaling behavior for the different chain lengths could not be deduced, it seems that the amino acid distance distributions can be attributed to geometric constraints of protein chains in bulk and amino acid sequences only play a secondary role.

\end{abstract}
\maketitle

\section{Introduction}
Proteins, the molecular machines of every living organism, perform vital tasks required for life to persist. These range from transport (e.g. hemoglobin)~\cite{ahmed2020hemoglobin}, signal transduction (e.g. rhodopsin)~\cite{nagata2021rhodopsins}, immune responses (e.g. antibodies), and hormonal regulation (e.g. insulin)~\cite{dill2008protein,Dill1042}. All natural proteins are made of 20 different amino acids which dictate the 3D conformations the proteins adopt in order to function~\cite{scheraga2007proteinfolding}. One of the great challenges has been to understand how the primary structure, i.e. the amino acid sequence, can lead to the fully folded functional protein, known as the protein folding problem~\cite{nassar2021protein}. The last 50 years have seen many different routes to computationally predict biologically active protein structures without having to solve a crystal structure of the protein~\cite{marks2012protein}. AlphaFold~2, a machine learning model for structure prediction, has shown predictiveness at a level of experimental accuracy in the most recent protein structure prediction challenge~\cite{jumper2021highly, kryshtafovych2021critical}. It was trained on structural data from the Research Collaboratory for Structural Bioinformatics Protein Data Bank~\cite{berman2000protein} (RCSB PDB). The RCSB PDB contains structures from crystallographic, NMR, and cryo-EM experiments. AlphaFold~2 certainly was a game changer in protein structure prediction~\cite{callaway2022what}, but it has not solved the protein folding problem in the sense that the mechanisms of the folding or structure function relationship remain unexplained. However, it does provide a rich resource of structures to understand emergent patterns in proteins on a more fundamental level.

One way of mapping the functional forms of proteins and classifying their structural properties is by looking at protein contact maps (PCM)~\cite{Vendruscolo2002,dipaola2013protein,Estrada2011}. A PCM is a matrix representation of spatially close amino acids, given a certain cut-off distance as determined typically from a protein structure. PCMs have shown promise in understanding protein folding patterns, as well as revealing allosteric communication pathways~\cite{yao2019establishing, menichetti2016network,dokholyan2002topological}. Their use is also prominently featured in machine learning approaches. For example, similarly to AlphaFold~2, one challenge is to predict a PCM \textit{de novo} from sequence alone~\cite{bassot2019using, rives2021biological, rao2021msa}, without using structural data as part of the training set. Another is the use PCMs in machine learning approaches to e.g. train a model able to predict binding affinities between a protein and a drug-like molecule~\cite{jiang2020drug}.

From a physicist's perspective, it is interesting to understand the emergent behavior, in terms of folding or function, in many protein structures. Therefore studying properties of underlying features of a PCM can reveal information on the scale of the proteome and not just individual proteins. Thus, having a reliable model to generate a PCM not from real data is invaluable. To this end, different heuristic models have been indtroduced in the past by Atilgan \textit{et al.}~\cite{atilgan2004smallworld} and Bartoli \textit{et al.}~\cite{bartoli2008effecta}. Atilgan \textit{et al.} propose a random shuffle model to generate artificial PCMs with similar properties to real protein models. Bartoli \textit{et al.} built a model for PCMs in which they assume ~\textit{ad hoc} that proteins have a distribution of amino acid distances $P(s)$ that follows $P(s) \approx s^{-1}$. The \textit{amino acid distance} $s$ is the separation of two connected amino acids along the backbone chain, and gives rise to the simplest implementation of a connection probability in a protein. Bartoli \textit{et al.} justify their approximation only heuristically in the sense that resulting PCMs have similar properties to real world PCMs.

In this paper, we provide an explanation why the $s^{-1}$ heuristic is a good initial assumption. To do so, we derived an analytical approximation from a 2D geometrically constrained protein model to get an expression for $P(s)$. The approximation provides additional correction terms. We then show that the approximation can be used to fit simulations of the 2D contstrained model~\cite{molkenthin2016scaling}, a 3D version of the constrained model~\cite{molkenthin2020self} and even to real world proteins. This means that the analytical approximation is a good starting point for artificially modelling PCMs in the future.

%We find that, despite the highly simplified 2D approximation and 2D and 3D models, the generated amino acid distance distribution $P(s)$ is in agreement with the one observed in PCMs generated from structural data and roughly consistent with the heuristic of $s^{-1}$. From the derived analytical approximation a correction term for the $s^{-1}$ heuristic, can be derived.

The paper is structured in the following way: In Sec.~\ref{sec:models}, we give a brief overview of the geometrically constrained models used, In Sec.~\ref{sec:methods}, we describe the construction of PCMs from the 2- and 3D simulations of the geometrically constrained model, as well as from structural data in the RCSB PDB and AlphaFold2 databases. In Sec.~\ref{sec:theory} we then derive an analytical approximation for the 2D analogue of the geometrically constrained model. Sec.~\ref{sec:results} highlights the results for the 2D, and 3D simulations, RCSB PDB, and AlphaFold~2 structural data by making use of the 2D analytical approximation.
%%%%%%%%%%%%%%%%%
% Section  %
%%%%%%%%%%%%%%%%%
\section{2D and 3D geometrically constrained models}
\label{sec:models}
Different approaches have been used in the past for geometrical models describing protein folding, some of which derived characteristics of the secondary structure from constraints on bond and torsion angles~\cite{bhattacharjee2013flory,Danielsson2010,Molkenthin2011} and others focused on the formation mechanisms of the tertiary structure~\cite{molkenthin2016scaling, molkenthin2020self} and even others are modelled by self avoiding random walks~\cite{mey2014rareevent, hills2009insights}. Here we will focus on models representing maximally folded structures in 2D, and 3D as introduced in~\cite{molkenthin2016scaling} and~\cite{molkenthin2020self} respectively. The 3D geometrical model and the analytical approximation of its 2D analogue both build on the idea that inherently geometrical objects, such as amino acids, imply that any resulting PCM network ensemble modeling them has to be spatially embedded.

In~\cite{molkenthin2016scaling}, we have introduced a simplified, 2D version of this geometrically constrained model. It starts from a closed chain (i.e. a ring) of $L$ unit discs and subsequently adds links, such that connected discs touch, yet no discs intersect. The advantage of this model is that it can be approximated by a purely topological simplification, which can be treated analytically. Fig.~\ref{fig:schematic} illustrates the construction of the 2D geometric model.

The topological constraints in the resulting network model are:
(a) New links always form between two units that are part of the same face of the graph (region enclosed by a cycle in the network). This prevents the overlapping of discs. (b) No links form across the outer face. This prevents the enclosure of a unit by less than six other units (which is geometrically impossible) such that (c) the maximum degree of each unit is six, as six is the maximum number of unit discs one central unit disc can touch. (d) Once connected by a link, pairs of units do not disconnect. In~\cite{molkenthin2020self} this geometrically constrained model was extended to 3D spheres using simulations to generate compact artificial protein structures. 
%%%%%%%%%%%%%%%%%
% Section  %
%%%%%%%%%%%%%%%%%
\section{Protein contact maps}
\label{sec:methods}
The complex interaction pattern between the amino acids in a protein can be naturally expressed as a network or graph, in which each amino acid is represented by a node and spatial proximity is encoded as a link. Whenever two central $\mathrm{C}_\alpha$ atoms are closer together than a threshold $d_c$, they are connected, linked, or in `close contact'. These connections can be determined from the 3D functional or folded structure of the protein and are encoded in a so-called (protein) contact map. This contact map is effectively an adjacency matrix $\textbf{A}^{\mathrm{struc}}$ 
\begin{equation}
  A^{\mathrm{struc}}_{ij}=
  \begin{cases}
   0, & \text{ if } d_{i,j}>d_c \text{ or } i=j\\
      1, & \text{ if } d_{i,j}\leq d_c.
      \end{cases}
    \label{eq:aij}
\end{equation}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\columnwidth]{figures/Fig1/Fig1.pdf}
    \caption{a) The amino acid distance $s$ (green) is defined as the number of amino acids (AS) (here represented by a circle) along the protein backbone chain (black) between two connected AS (pink). b) Shows an example AS distance of length $s=3$ (solid green line). When adding a new connection from node A, the previously created connection prohibits new ones (dashed pink lines). The green dotted link from A is still permitted and can be made. c) Shows an example of an existing AS distance $s=7$ (solid green), where there are now more permitted (dotted green) links and fewer prohibited (dotted pink) ones.}
    \label{fig:schematic}
\end{figure}

@@

Based on the simple observation that amino acids are objects in space which cannot overlap indefinitely, we introduced the 3D version of the 2D geometrically constrained model in~\cite{molkenthin2020self}. Starting from a chain of identical spheres, each in contact only with the neighbors it is connected to, additional connections are made by randomly selecting two spheres and moving them towards each other until they touch. The new links formed that way cannot be broken in later steps and function as constraints on subsequent link formation steps. If the contact of the two selected spheres is geometrically impossible without breaking previously made connections or leading to overlaps of any spheres, the link is not made and is taken out of the pool of possible connections. This process is repeated until no more links can be formed without violating the geometric constraints and the final structure can be expressed as the adjacency matrix
\begin{equation}
  A^{\mathrm{sim}}_{ij}=
  \begin{cases}
   1, & \text{ if }|i-j|=1 \text{ or i and j connected} \\
      0, & \text{ otherwise}.
      \end{cases}
    \label{eq:sim_aij}
\end{equation}

%In~\cite{molkenthin2016scaling}, we have introduced a simplified, two-dimensional version of this model, which starts from a closed chain (i.e. a ring) of $L$ unit discs and subsequently adds links, such that connected discs touch, yet no discs intersect. The advantage of this model is that it can be approximated by a purely topological simplification, which can be treated analytically. Fig.~\ref{fig:schematic} illustrates the construction of the 2D geometric model.

%The topological constraints in the resulting network model are:
%(a) New links always form between two units that are part of the same face of the graph (region enclosed by a cycle in the network). This prevents the overlapping of discs. (b) No links form across the outer face. This prevents the enclosure of a unit by less than six other units (which is geometrically impossible) such that (c) the maximum degree of each unit is six, as six is the maximum number of unit discs one central unit disc can touch. (d) Once connected by a link, pairs of units do not disconnect.

@@

The underlying formation mechanisms of individual protein folds are incredibly complex and depend on the surrounding solvent, as well as the specific amino acid sequence and their quantum-mechanical interactions. Here we want to assess if the geometrically constrained protein model's contact map distributions capture real protein behavior and make it a viable model for PCMs and provide a better explanation for the heuristic used in Bartoli \textit{et al.}~\cite{bartoli2008effecta}. To understand this better, we study the ensemble of PCMs from real protein structures and will look at `averaged' contact maps over many different proteins that have different amino acid sequences, but their overall sequence or chain length is the same. 
We will use the RCSB Protein Data Bank (RCSB PDB), a database of structurally resolved protein amino acid sequences through X-ray crystallography, NMR or cryo-EM experiments and the synthetically generated structural database from the machine learned structures from the AlphaFold~2 database~\cite{varadi2022alphafold}. The RCSB PDB contains just under 200,000 protein structures to date~\cite{berman2000protein}. The distribution of protein chain lengths in the RCSB PDB is illustrated in Fig.~\ref{fig:pdb_stats} b. These chain lengths vary from small fragments of less than 10 amino acids to large agglomerates of over 2000 amino acids. The most frequently occurring chain lengths, however, are between 100 and 300 amino acids as shown in Fig.~\ref{fig:pdb_stats} b. The pink subset of the bar plot shown in Fig.~\ref{fig:pdb_stats} b, is the set of structures used that fulfill the criterion on protein chain lengths of $L\approx100$ ($85\le L\le 115$) amino acids, $L\approx 200$ ($185 \le L \le 215$) amino acids and $L\approx300$ ($285 \le L\le 315$) amino acids and represent all RCSB PDB structures used in our analysis. The second database consists of synthetically generated structures using AlphaFold~2 with sequences taken from SwissProt~\cite{bairoch2000swissprot}. SwissProt is a manually curated database of protein sequences containing over 500,000 protein sequences, whose protein chain length distribution is shown in Fig.~\ref{fig:pdb_stats} a in teal. The pink bars indicate the subset of AlphaFold~2 structures used for the contact map analysis. In this case, the initial structures were selected based on the condition of chain lengths in the same three intervals as used for the RCSB PDB. Second, the AlphaFold~2 per-residue confidence scores~\cite{jumper2021highly} were used to select proteins that had an average score of 90 or higher. Lastly, we removed structures that described the same protein but originated from different organisms, by selecting unique protein names from the pandas data table.

 \begin{figure}[htb]
        \centering
	\includegraphics[width=0.9\columnwidth]{paper/figures/Fig2/Fig2.pdf}
	    \caption{Bar plot of protein sequence lengths for protein structures from two structure databases. a) AlphaFold~2 predicted protein structures taken from protein sequences in the manually curated SwissProt database of protein sequences. Pink bars show fraction of structures used for the average contact map analysis. b) RSCB PDB chain length frequencies. Pink bars show protein structures whose chain lengths matches the resolution of structure accounting for the whole sequence. Only pink RCSB PDB samples were used for the average contact map analysis. c) Example of a protein contact map taken from RCSB PDB ID: 3UFG, with a cut-off of 8~Å. Pink squares indicate a contact present between amino acids.}
        \label{fig:pdb_stats}
\end{figure}
All structure-based PCMs were generated in the same way for AlphaFold~2 structures and RCSB PDB structures. Structures used were downloaded in pdb format, read in with MDAnalysis version 2.0.0~\cite{gowers2016mdanalysis} and only C$_{\alpha}$ were selected and used for the analysis. Proteins were categorised by length and placed into the same three groups as for the RCSB PDB data. For each of the sequences, all pairwise C$_\alpha$ distances were calculated, and two atoms were said to be in contact if their distance was less than 8~Å. The contacts (1 or 0) were then recorded in the adjacency matrix $\mathbf{A}^{\mathrm{struc}}$. An example of a protein contact map, or adjacency matrix, is shown in Fig.~\ref{fig:pdb_stats}~c. In order to study the amino acid distance distributions of real and simulated protein structures, the distances were calculated from the adjacency matrices. These distances were then histogrammed to calculate the mean amino acid distance distributions shown in Sec.~\ref{sec:results}. The pseudoalgorithms describing the above two processes are shown in the Supplemental Material~(SI)~\cite{SI}. All code for the construction of the contact maps and their analysis is available on GitHub at~\cite{2022sequence}. The derived analytical approximation in Sec.~\ref{sec:theory}, was used for data fitting. All fits were carried out using Scipy 1.7.0 and with scripts for the fits available on GitHub at~\cite{2022sequence}.

%%%%%%%%%%%%
% Section. %
%%%%%%%%%%%%
\section{Analytical approximation for the geometrically constrained model}\label{sec:theory}

To assess how we can best describe the distribution of the probability of amino acid distances $P(s)$ obtained from a PCM for the geometrically constrained model, we consider the amino acid distance distribution for the simplified 2D model as proposed in~\cite{molkenthin2016scaling}. We are thus able to propose the following analytical approximation. Fig~\ref{fig:schematic} summarizes the idea behind this two-dimensional model.

Let us start by introducing the auxiliary variable $F_k(s)$, defined to be the number of possible links with an amino acid distance of $s$ if $k$ links have been added before. Before any links are added, all amino acid distances are equally likely, as there are $F_0(s)=N$ possibilities of making a link of distance $s$ each, with $2\leq s < \frac{N}{2}$.

As we add more links, not only are links taken out of this pool, because they have already been realized, but each existing link can also geometrically prohibit other connections. Adding a link of length $s_1$ prohibits $2(s-1)$ links of length $s<s_1$ and $2(s_1-1)$ links of length $s\geq s_1$ (see Fig.~\ref{fig:schematic}~b). To find the expected number of links still available in subsequent steps, we average over all possible amino acid distances $s_1$ of the first link added.

The expected distribution of possible links after one step is thus given by the average over available link pools taken over all possible values of $s_1$:
